Safe Transferable Regions

### 2015-11-20 paper/abstract.tex
small things in intro.

Recent work in high-performance systems written in managed languages
(such as Java or C\#) has shown that garbage-collection can be a
significant performance bottleneck in such systems.  A class of these
systems, focused on big-data, create huge data-structures with
well-defined lifetimes.  In this paper, we present a language and
memory management scheme based on user-managed memory regions (called
\emph{transferable regions}) that allow programmers to exploit
knowledge of data-structures' lifetimes to achieve significant
performance improvements.

>>>>

Recent work in high-performance 
systems written in managed languages 
(such as Java or C\#) has shown that garbage-collection can be a
significant performance bottleneck.  A class of these
systems, focused on big-data, create many and often large data-structures with
well-defined lifetimes.  In this paper, we present a language and
memory management scheme based on user-managed memory regions (called
\emph{transferable regions}) that allow programmers to exploit
knowledge of data-structures' lifetimes to achieve significant
performance improvements.

### 2015-11-20 paper/introduction.tex
small things in intro.

This example is an instance of a general pattern, where a producer
creates a data-structure and passes it to a consumer. In a system
where most of the computation takes this form, and these
data-structures are very large, as is the case with many streaming
big-data analysis systems, garbage collection overhead becomes
significant~\cite{Broom:HotOS}.
Furthermore, in a distributed dataflow system, the GC pause at one node can
have a cascading adverse effect on the performance of other nodes.
For instance, a GC pause at an upstream actor can block
downstream actors that are waiting for messages. 
Much of the GC overhead results from the collector performing avoidable or unproductive work.
For example, in the process executing the code from
Fig.~\ref{fig:motivating-eg}, GC might repeatedly traverse the \C{map}
data-structure, although its objects cannot be collected until a
suitable timing message arrives.

@@

An important observation, in the context of processes of kind
described above, is that the data-structures exchanged between them
can be partitioned into sets of fate-sharing objects with common
lifetimes, which makes them good candidates for a region-based memory
management discipline. A region is a block of memory that is allocated
and freed in one shot, consuming constant time. A region may contain
one or more contiguous range of memory locations, and individual
objects may be dynamically allocated within the region over time,
while they are deallocated en masse when the region is freed.  Thus, a
region is a good fit for a set of fate-sharing objects.
% that is used to realize a data-structure with a well-defined
% lifetime. 
In this example, the output to be constructed for each time window $t$
can be implemented using a separate region.

@@

This ensures that we can safely free one dynamic region (\eg, the
input-message in our example) without affecting the validity of
pointers in another dynamic region (\eg, the output-message).
However, this is not sufficient! When processing the input-message
$D$, we necessarily will have to create pointers that point to objects
that are internal to $D$. Such pointers may be stack-allocated or even
reside in the heap (\eg, consider the iterator object used to iterate
over the list in the input-message).  How do we ensure the safety of
these pointers?

>>>>

This example is an instance of a general pattern, where a producer
creates a data-structure and passes it to a consumer. In a system
where most of the computation takes this form, and these
data-structures are very large, as is the case with many streaming
big-data analysis systems, garbage collection overhead becomes
significant~\cite{Broom:HotOS}.
Furthermore, in a distributed dataflow system, the GC pause at one node can
have a cascading adverse effect on the performance of other nodes, particularly
when real-time streaming performance is required.
\dv{citation missing here, Tim Haris's work and our HotOS submission.}
In particular, a GC pause at an upstream actor can block
downstream actors that are waiting for messages. 
However, often much of the GC overhead results from the collector performing avoidable or unproductive work.
For example, in the process executing the code from
Fig.~\ref{fig:motivating-eg}, GC might repeatedly traverse the \C{map}
data-structure, although its objects cannot be collected until a
suitable timing message arrives.

@@

An important observation, in the context of processes of the kind
described above, is that the data-structures exchanged between them
can be partitioned into sets of fate-sharing objects with common
lifetimes, which makes them good candidates for a region-based memory
management discipline. A region is a block of memory that is allocated
and freed in one shot, consuming constant time. A region may contain
one or more contiguous range of memory locations, and individual
objects may be dynamically allocated within the region over time,
while they are deallocated en masse when the region is freed.  Thus, a
region is a good fit for a set of fate-sharing objects.
% that is used to realize a data-structure with a well-defined
% lifetime. 
In this example, the output to be constructed for each time window $t$
can be implemented using a separate region.

@@

This ensures that we can safely free one dynamic region (\eg, the
input-message in our example) without affecting the validity of
pointers in another dynamic region (\eg, the output-message).
However, this is not sufficient! \dv{for what} When processing the input-message
$D$, we necessarily will have to create pointers that point to objects
that are internal to $D$. Such pointers may be stack-allocated or even
reside in the heap (\eg, consider the iterator object used to iterate
over the list in the input-message).  How do we ensure the safety of
these pointers?
